Projecting genetic associations through gene expression patterns highlights disease etiology and drug mechanisms

### 2023-06-13 output/latex/manuscript.tex
methods: fix and remove reference to non existing supplementary file

We found that phenotypes were grouped into five clear branches, defined by their first node at the top of the Figure \ref{fig:clustering:tree}:
0) a ``large'' branch that includes most of the traits subdivided only starting at \(k\)=16 (with asthma, subjective well-being traits, and nutrient intake clusters),
1) heel bone-densitometry measurements,
2) hematological assays on red blood cells,
3) physical measures, including spirometry and body impedance, and anthropometric traits with fat-free and fat mass measures in separate sub-branches, and
4) a ``complex'' branch including keratometry measurements, assays on white blood cells and platelets, skin and hair color traits, autoimmune disorders, and cardiovascular diseases (which also included other cardiovascular-related traits such as hand-grip strength\textsuperscript{\protect\hyperlink{ref-aBVh8zt1}{81}}, and environmental/behavioral factors such as physical activity and diet) (see Supplementary Files 3-6 for clustering results).
Within these branches, results were relatively stable, with the same traits often clustered together across different resolutions.
Arrows between clusters show traits moving from one group to another, and this mainly happens between clusters within the ``complex'' branch (4) and between clusters from the ``large'' branch (0) to the ``complex'' branch.
This behavior is expected since complex diseases are usually associated with shared genetic and environmental factors and are thus hard to categorize into a single cluster.

@@

We used two large TWAS resources from different cohorts for discovery and replication, all obtained from European ancestries.
PhenomeXcan\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42}}, our discovery cohort, provides results on 4,091 traits across different categories.
Supplemenetary File 1 has all the details about the included GWAS, sample size and disease/trait categories.
In PhenomeXcan, these publicly available GWAS summary statistics were used to compute
1) gene-based associations with the PrediXcan family of methods (described before), and
2) a posterior probability of colocalization between GWAS loci and \emph{cis}-eQTL with fastENLOC\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42},\protect\hyperlink{ref-ndd3tW4g}{96}}.
We refer to the matrix of \(z\)-scores from S-PrediXcan (Equation (\ref{eq:spredixcan})) across \(q\) traits and \(m\) genes in tissue \(t\) as \(\mathbf{M}^{t} \in \mathbb{R}^{q \times m}\).
As explained later, matrices \(\mathbf{M}^{t}\) were used in our LV-based drug repurposing framework since they provide direction of effects.
The S-MultiXcan results (22,515 gene associations across 4,091 traits) were used in our LV-based regression framework and our cluster analyses of traits.
For the cluster analyses, we used the \(p\)-values converted to \(z\)-scores: \(\mathbf{M}=\Phi^{-1}(1 - p/2)\), where \(\Phi^{-1}\) is the probit function.
Higher \(z\)-scores correspond to stronger associations.

>>>>

We found that phenotypes were grouped into five clear branches, defined by their first node at the top of the Figure \ref{fig:clustering:tree}:
0) a ``large'' branch that includes most of the traits subdivided only starting at \(k\)=16 (with asthma, subjective well-being traits, and nutrient intake clusters),
1) heel bone-densitometry measurements,
2) hematological assays on red blood cells,
3) physical measures, including spirometry and body impedance, and anthropometric traits with fat-free and fat mass measures in separate sub-branches, and
4) a ``complex'' branch including keratometry measurements, assays on white blood cells and platelets, skin and hair color traits, autoimmune disorders, and cardiovascular diseases (which also included other cardiovascular-related traits such as hand-grip strength\textsuperscript{\protect\hyperlink{ref-aBVh8zt1}{81}}, and environmental/behavioral factors such as physical activity and diet) (see Supplementary Files 2-6 for clustering results).
Within these branches, results were relatively stable, with the same traits often clustered together across different resolutions.
Arrows between clusters show traits moving from one group to another, and this mainly happens between clusters within the ``complex'' branch (4) and between clusters from the ``large'' branch (0) to the ``complex'' branch.
This behavior is expected since complex diseases are usually associated with shared genetic and environmental factors and are thus hard to categorize into a single cluster.

@@

We used two large TWAS resources from different cohorts for discovery and replication, all obtained from European ancestries.
PhenomeXcan\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42}}, our discovery cohort, provides results on 4,091 traits across different categories.
In PhenomeXcan, these publicly available GWAS summary statistics were used to compute
1) gene-based associations with the PrediXcan family of methods (described before), and
2) a posterior probability of colocalization between GWAS loci and \emph{cis}-eQTL with fastENLOC\textsuperscript{\protect\hyperlink{ref-lY5ln3dB}{42},\protect\hyperlink{ref-ndd3tW4g}{96}}.
We refer to the matrix of \(z\)-scores from S-PrediXcan (Equation (\ref{eq:spredixcan})) across \(q\) traits and \(m\) genes in tissue \(t\) as \(\mathbf{M}^{t} \in \mathbb{R}^{q \times m}\).
As explained later, matrices \(\mathbf{M}^{t}\) were used in our LV-based drug repurposing framework since they provide direction of effects.
The S-MultiXcan results (22,515 gene associations across 4,091 traits) were used in our LV-based regression framework and our cluster analyses of traits.
For the cluster analyses, we used the \(p\)-values converted to \(z\)-scores: \(\mathbf{M}=\Phi^{-1}(1 - p/2)\), where \(\Phi^{-1}\) is the probit function.
Higher \(z\)-scores correspond to stronger associations.
